Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-06 _paper.rmd
done with SHAP and local explanations

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001 and @shmueli_explain_2010 introduce the idea of distinguishing modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample observations. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

<!-- Interpretability & baises -->
Interpretability is vital for exploring and protecting against potential biases in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response variable, which is then built into the model. Such biases include sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]. Variable-level interpretability of models is essential in the evaluation models for such biases.

<!-- Interpretability & data drift -->
Another concern is that of data drift, which is a shift in support or domain of the explanatory variables. Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

<!-- Local explanations -->
Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use _local explanations_, which attempt to approximate linear variable importance at the location of each observation, or the predictions at a specific point in the data domain. Because these are point specific, a challenge is to visualize to more comprehensively understand a model.

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanation statistics](#sec:explanations), covers the background of the local explanation, and the traditional visuals produced. [Tours and the radial tour](#sec:tour) explains the animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with a [Discussion](#sec:cheemdiscussion) of the insights gained and directions that might be explore in the future. (XXX DC Can you change the references to number of the section not title?)

# Local explanation statistics {#sec:explanations}

<!-- Reminder of local explanation -->
Consider a highly non-linear model. At face value, it is hard to say which variable(s) are sensitive to  crossing a classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

<!-- Taxonomy of local explanations -->
Figure six of @arrieta_explainable_2020 gives comprehensive summarization of the taxonomy and literature of explanation techniques. The figure includes a large number of model-specific explanations such as deepLIFT, [@shrikumar_not_2016; @shrikumar_learning_2017] a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME, [@ribeiro_why_2016] SHAP [@lundberg_unified_2017] and their variants are popular.

<!-- Uses of local explanations -->
These instance-level explanations are used in various ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating necessary pixels [@simonyan_deep_2014]. For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction [@ribeiro_why_2016].

## SHAP and tree SHAP

<!-- SHAP and history -->
SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation by taking the median importance of a subset of permutations in the explanatory variables. This idea stems from the field of game theory, where @shapley_value_1953 devised a method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score.

<!-- Fifa example -->
To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season [@leone_fifa_2020]. We have 5000 observations of nine skill measures (after aggregating highly correlated variables). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

```{r shapdistrbd, echo=F, fig.align='center', fig.cap = "Illustration of the distribution of SHAP attributions and a break-down plot. From FIFA 2020 data, a random forest model regresses wages from nine skill attributes for a star offensive and defensive player. The players have very different salaries, but a) shows the distributions of 25 permutations in the explanatory variables. The medians of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances seem realistic given the player's positions. b) Break-down plots of the observations using their explanations to additively cover the difference between the model intercept and the observation predictions."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

<!-- Illustration -->
Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offense and movement skills are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel b) shows a simplified break-down plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observation's prediction. Such additive approaches will show asymmetry in the variable ordering, so we opt to fix the order to panel a) by decreasing the sum of the SHAP values.

<!-- Segue -->
In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application, we apply _tree SHAP_, a variant of SHAP enjoys a lower computational complexity [@lundberg_consistent_2018]. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis to further scrutinize the explanation.

@@

## Manual tours and its radial case

@@

# cheem viewer {#sec:applicationdesign}

>>>>

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001, reiterated by @shmueli_explain_2010, taxonomizes modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample observations. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

<!-- Interpretability & baises -->
Interpretability is vital for exploring and protecting against potential biases (e.g. sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) variable, which is then built into the model. Variable-level (feature-level) interpretability of models is essential in the evaluation models for such biases. It is also generally important for many problems, where it is not enough to accurately predict accurately but one must be able to explain which predictors are most responsible in generating a response value.

<!-- Interpretability & data drift -->
Another concern is that of data drift, which is a shift in support or domain of the explanatory variables (feature or predictors). Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

<!-- Local explanations -->
Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use _local explanations_, which attempt to approximate linear variable importance at the location of each observation (instance), or the predictions at a specific point in the data domain. Because these are point specific, a challenge is visualizing them to more comprehensively understand a model.

XXX DC Note that we need to make sure that the paper is readable for a machine learner, so a translation of terms is important. I've done this by parenthesizing the ML term in this introduction, which should be enough.

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanations](#sec:explanations) covers the background of the local explanation, and the traditional visuals produced. [Tours and the radial tour](#sec:tour) explains the animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with a [Discussion](#sec:cheemdiscussion) of the insights gained and directions that might be explore in the future. (XXX DC Can you change the references to number of the section not title?)

# Local explanations {#sec:explanations}

<!-- Reminder of local explanation -->
Consider a highly non-linear model. It can be hard to determine whether small changes in a variable's value will make a class prediction change group or identify which variables contribute to an extreme residual. Local explanations shed light on these situations by approximating linear variable importance in the vicinity of a single observation or point.

<!-- Taxonomy of local explanations -->
A comprehensive summary of the taxonomy and literature of explanation techniques is provided in Figure 6 of @arrieta_explainable_2020. It includes a large number of model-specific explanations such as deepLIFT [@shrikumar_not_2016; @shrikumar_learning_2017], a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME [@ribeiro_why_2016] SHAP [@lundberg_unified_2017], and their variants are popular.

<!-- Uses of local explanations -->
These instance-level explanations are used in various ways depending on the data. In image classification, where pixels would correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels [@simonyan_deep_2014]. For instance, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction [@ribeiro_why_2016].

<!-- ## SHAP SHAP and history -->
SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. However, it is also related to partial dependence plots (XXX Needs reference), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each variable in the context of all other candidate variables.

For our application, we use _tree SHAP_, a variant of SHAP enjoys a lower computational complexity [@lundberg_consistent_2018]. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis to further scrutinize the explanation.

<!-- Fifa example -->
XXX DC does this example come from/motivated by https://ema.drwhy.ai/UseCaseFIFA.html ?? it needs to be cited here. 
To understand the use SHAP we use soccer data from FIFA 2020 season [@leone_fifa_2020]. There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit with log wages, in 2020 Euros, as the response, using the skill measures as predictors. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where the order of variables is fixed, so the two observations can be more easily compared.

```{r shapdistrbd, echo=F, fig.align='center', fig.cap = "Illustration of the use of SHAP values for a random forest model for salaries for FIFA 2020 players based on nine predictors corresponding to different skills. Two observations (a star offensive player, Messi and a defensive player, van Dijk) are compared. Plot (a) shows the distribution of contributions for each variable across 25 permutations of predictors, with the median displayed as a dot, for both players. Plot b) is the break-down plot showing the SHAP values. It can be learned that reaction is important for explaining both players salaries. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

XXX DC Can you remove "Variable" from the vertical axes? And remove reference to "permutations" in the figure text. Also, why include interaction and prediction - suggest removing these because there is no information. Perhaps switch the order, and put the breakdown plot first?

<!-- Illustration -->
<!-- Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offense and movement skills are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel b) shows a simplified break-down plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observation's prediction. Such additive approaches will show asymmetry in the variable ordering, so we opt to fix the order to panel a) by decreasing the sum of the SHAP values.-->

<!-- Segue -->
In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages.

@@

<!-- ## Manual tours and its radial case-->

@@

# The cheem viewer {#sec:applicationdesign}
